---
title: "Рисунки для статьи"
author: ""
date: ''
output: html_document
---




```{r setup, include=FALSE}
library(knitr)
  
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(ggplot2)
library(mgcv)
library(dplyr)
library(readxl)
library(cowplot)


theme_set(theme_bw())
```


```{r}

finch_il <- read_excel("Data/Spinus_2020_parasitemia_IL6_survival_final.xlsx", na = "NA", sheet = "Spinus_2020_parasitemia_IL6_sur")
finch_il$Experiment <- factor(finch_il$Experiment, labels = c("Control", "SGS1", "GRW2"))
finch_il$Ring <- factor(finch_il$Ring)

n_obs_il <- finch_il %>% group_by(Ring) %>% summarise(N_il_obs = n())


finch <- read_excel("Data/Spinus_2020_parasitemia_IL6_survival_final.xlsx", na = "NA", sheet = "RMR")
finch$Experiment <- factor(finch$Experiment, labels = c("Control", "SGS1", "GRW2"))
finch$Ring <- factor(finch$Ring)

# n_obs_rmr <- finch %>% group_by(Ring) %>% summarise(N_rmr_obs = n())
# 
# n_obs <- merge(n_obs_il, n_obs_rmr, by = "Ring")

finch_il <- finch_il %>% group_by(Ring) %>% arrange(Date) %>% mutate(Cum_Parasitemia = cumsum(Parasitemia), Cum_IL = cumsum(IL6_pg_ml)) %>% arrange(Ring, Date)


```




```{r}
# Вычисление паразитемии по двум соседним точкам измерения от дня измерения RMR  



df_rmr <- finch %>%  select(Ring, Date, Experiment, RMR) %>% mutate(Type = "RMR", Value = RMR) %>% select(-RMR)

df_par <- finch_il %>% select(Ring, Date, Experiment,Parasitemia) %>% mutate(Type = "Parasitemia", Value = Parasitemia) %>% select(-Parasitemia)

df_all <- rbind(df_rmr, df_par)

df_all <- df_all %>% group_by(Experiment, Ring) %>% arrange(Date, .by_group = TRUE)   

df_all_check <- df_all %>% group_by(Ring) %>% summarise(RMR_count = sum(Type == "RMR"))

# Удаляем птиц, у которых не измеряли RMR
exclude_Ring <- df_all_check %>% filter(RMR_count == 0) %>% pull(Ring)

df_all <- df_all %>% filter(!Ring %in% exclude_Ring)


df_final <- NULL

for(Selected_Ring in unique(df_all$Ring)){
  df <- df_all %>% filter(Ring == Selected_Ring)
  df <- df %>% arrange(Date)
  
  k <- 1
  df_estim <- data.frame(Ring = rep(NA, nrow(df)), Date = df$Date[1], Experiment = NA, RMR = NA, Parasitemia = NA )
  
  for(i in 1:nrow(df)){
    if(df$Type[i] == "RMR"){
      
      paras <- df[c(i-1, i+1), ]
      
      if(nrow(paras) == 1){
      df_estim$Ring[k] <- as.character(Selected_Ring)
      df_estim$RMR [k] <- df$Value[i]
      df_estim$Date [k] <- (df$Date[i])
      df_estim$Experiment [k] <- as.character(df$Experiment[i])
      df_estim$Parasitemia [k] <- 0  
      }
      else
      {
      mod <- lm(Value ~ Date, data = paras)
      df_estim$Ring[k] <- as.character(Selected_Ring)
      df_estim$RMR [k] <- df$Value[i]
      df_estim$Date [k] <- (df$Date[i])
      df_estim$Experiment [k] <- as.character(df$Experiment[i])
      df_estim$Parasitemia [k] <- predict(mod, newdata = df[i,])  
      }
      k <- k + 1
    }
  }
  
  df_estim <- df_estim %>% filter(complete.cases(.)) 

  df_final <- rbind(df_final, df_estim)  
  
}




```




```{r}
# Вычисление Интерлейкина по двум соседним точкам измерения от дня измерения RMR, для первой точки измерения RMR берется первое измерение интерлейкина, следующее после измерения RMR  



df_rmr <- finch %>%  select(Ring, Date, Experiment, RMR) %>% mutate(Type = "RMR", Value = RMR) %>% select(-RMR)

df_il <- finch_il %>% select(Ring, Date, Experiment,IL6_pg_ml) %>% mutate(Type = "IL", Value = IL6_pg_ml) %>% select(-IL6_pg_ml)

df_all2 <- rbind(df_rmr, df_il)

df_all2 <- df_all2 %>% group_by(Experiment, Ring) %>% arrange(Date, .by_group = TRUE)   

df_all2_check <- df_all2 %>% group_by(Ring) %>% summarise(RMR_count = sum(Type == "RMR"))

# Удаляем птиц, у которых не измеряли RMR
exclude_Ring2 <- df_all2_check %>% filter(RMR_count == 0) %>% pull(Ring)

df_all2 <- df_all2 %>% filter(!Ring %in% exclude_Ring2)


df_final2 <- NULL

for(Selected_Ring in unique(df_all2$Ring)){
  df <- df_all2 %>% filter(Ring == Selected_Ring)
  df <- df %>% arrange(Date)
  
  k <- 1
  df_estim <- data.frame(Ring = rep(NA, nrow(df)), Date = df$Date[1], Experiment = NA, RMR = NA, IL = NA )
  
  for(i in 1:nrow(df)){
    if(df$Type[i] == "RMR"){
      
      IL <- df[c(i-1, i+1), ]
      
      if(nrow(IL) == 1){
      df_estim$Ring[k] <- as.character(Selected_Ring)
      df_estim$RMR [k] <- df$Value[i]
      df_estim$Date [k] <- as.POSIXct(df$Date[i])
      df_estim$Experiment [k] <- as.character(df$Experiment[i])
      df_estim$IL [k] <- df$Value[i+1]  
      }
      else
      {
      mod <- lm(Value ~ Date, data = IL)
      df_estim$Ring[k] <- as.character(Selected_Ring)
      df_estim$RMR [k] <- df$Value[i]
      df_estim$Date [k] <- as.POSIXct(df$Date[i])
      df_estim$Experiment [k] <- as.character(df$Experiment[i])
      df_estim$IL [k] <- predict(mod, newdata = df[i,])  
      }
      k <- k + 1
    }
  }
  
  df_estim <- df_estim %>% filter(complete.cases(.)) 

  df_final2 <- rbind(df_final2, df_estim)  
  
}




```




```{r}
# Вычисление накопленной паразитемии по двум соседним точкам измерения от дня измерения RMR  



df_rmr <- finch %>%  select(Ring, Date, Experiment, RMR) %>% mutate(Type = "RMR", Value = RMR) %>% select(-RMR)

df_cum_par <- finch_il %>% select(Ring, Date, Experiment,Cum_Parasitemia) %>% mutate(Type = "Cum_Parasitemia", Value = Cum_Parasitemia) %>% select(-Cum_Parasitemia)

df_all3 <- rbind(df_rmr, df_cum_par)

df_all3 <- df_all3 %>% group_by(Experiment, Ring) %>% arrange(Date, .by_group = TRUE)   

df_all3_check <- df_all3 %>% group_by(Ring) %>% summarise(RMR_count = sum(Type == "RMR"))

# Удаляем птиц, у которых не измеряли RMR
exclude_Ring <- df_all3_check %>% filter(RMR_count == 0) %>% pull(Ring)

df_all3 <- df_all3 %>% filter(!Ring %in% exclude_Ring)


df_final3 <- NULL

for(Selected_Ring in unique(df_all3$Ring)){
  df <- df_all3 %>% filter(Ring == Selected_Ring)
  df <- df %>% arrange(Date)
  
  k <- 1
  df_estim <- data.frame(Ring = rep(NA, nrow(df)), Date = df$Date[1], Experiment = NA, RMR = NA, Cum_Parasitemia = NA )
  
  for(i in 1:nrow(df)){
    if(df$Type[i] == "RMR"){
      
      cum_paras <- df[c(i-1, i+1), ]
      
      if(nrow(cum_paras) == 1){
      df_estim$Ring[k] <- as.character(Selected_Ring)
      df_estim$RMR [k] <- df$Value[i]
      df_estim$Date [k] <- (df$Date[i])
      df_estim$Experiment [k] <- as.character(df$Experiment[i])
      df_estim$Cum_Parasitemia [k] <- 0  
      }
      else
      {
      mod <- lm(Value ~ Date, data = cum_paras)
      df_estim$Ring[k] <- as.character(Selected_Ring)
      df_estim$RMR [k] <- df$Value[i]
      df_estim$Date [k] <- (df$Date[i])
      df_estim$Experiment [k] <- as.character(df$Experiment[i])
      df_estim$Cum_Parasitemia [k] <- predict(mod, newdata = df[i,])  
      }
      k <- k + 1
    }
  }
  
  df_estim <- df_estim %>% filter(complete.cases(.)) 

  df_final3 <- rbind(df_final3, df_estim)  
  
}




```





```{r}
# Вычисление накопленного инртерлейкина по двум соседним точкам измерения от дня измерения RMR  



df_rmr <- finch %>%  select(Ring, Date, Experiment, RMR) %>% mutate(Type = "RMR", Value = RMR) %>% select(-RMR)

df_cum_il <- finch_il %>% select(Ring, Date, Experiment,Cum_IL) %>% mutate(Type = "Cum_IL", Value = Cum_IL) %>% select(-Cum_IL)

df_all4 <- rbind(df_rmr, df_cum_il)

df_all4 <- df_all4 %>% group_by(Experiment, Ring) %>% arrange(Date, .by_group = TRUE)   

df_all4_check <- df_all4 %>% group_by(Ring) %>% summarise(RMR_count = sum(Type == "RMR"))

# Удаляем птиц, у которых не измеряли RMR
exclude_Ring <- df_all4_check %>% filter(RMR_count == 0) %>% pull(Ring)

df_all4 <- df_all4 %>% filter(!Ring %in% exclude_Ring)


# Удаляем даты, когда Cum_IL == NA

df_all4 <- df_all4 %>% filter(!is.na(Value))


df_final4 <- NULL

for(Selected_Ring in unique(df_all4$Ring)){
  df <- df_all4 %>% filter(Ring == Selected_Ring)
  df <- df %>% arrange(Date)
  
  k <- 1
  df_estim <- data.frame(Ring = rep(NA, nrow(df)), Date = df$Date[1], Experiment = NA, RMR = NA, Cum_IL = NA )
  
  for(i in 1:nrow(df)){
    if(df$Type[i] == "RMR"){
      
      cum_IL <- df[c(i-1, i+1), ]
      
      if(nrow(cum_IL) == 1){
      df_estim$Ring[k] <- as.character(Selected_Ring)
      df_estim$RMR [k] <- df$Value[i]
      df_estim$Date [k] <- (df$Date[i])
      df_estim$Experiment [k] <- as.character(df$Experiment[i])
      df_estim$Cum_IL [k] <- df$Value[i+1]
      }
      else
      {
      mod <- lm(Value ~ Date, data = cum_IL)
      df_estim$Ring[k] <- as.character(Selected_Ring)
      df_estim$RMR [k] <- df$Value[i]
      df_estim$Date [k] <- (df$Date[i])
      df_estim$Experiment [k] <- as.character(df$Experiment[i])
      df_estim$Cum_IL [k] <- predict(mod, newdata = df[i,])  
      }
      k <- k + 1
    }
  }
  
  df_estim <- df_estim %>% filter(complete.cases(.)) 

  df_final4 <- rbind(df_final4, df_estim)  
  
}




```





```{r}
# data.frame с оценками паразитемии и интерлейкина по соседним точкам
rmr_paras_il <- merge(df_final, df_final2) %>% merge(., df_final3) %>% merge(., df_final4) 

rmr_paras_il <- rmr_paras_il %>% group_by(Ring, Experiment) %>% arrange(Date, .by_group = T)


```





```{r}
D_RMR <- finch %>% group_by(Ring) %>% summarise(D_RMR = log(RMR) - log(RMR)[1]) %>% pull


finch$D_RMR <- D_RMR


# ggplot(finch, aes(x = DPI, y = D_RMR , group = Ring)) + geom_line() + facet_wrap(~Experiment)

```






```{r}

# Код для совмещения данных по RMR и измерениям паразитемии и интерлейкинов
finch$Parasitemia_nearest <- NA
finch$Parasitemia_M <- NA

finch$IL_nearest <- NA
finch$IL_M <- NA

finch$Cum_Parasitemia_nearest <- NA
finch$Cum_Parasitemia_M <- NA

finch$Dif_Date <- NA
finch$Survival <- NA

for(i in 1:nrow(finch)){
  df <- finch_il %>% filter(Ring %in% finch$Ring[i])

  dif_dat1 <- (difftime(df$Date, finch$Date[i]))
  dif_dat1[dif_dat1 < 0 ] <- NA
  time1 <- df$Date[which.min(dif_dat1)]
  
  finch$Parasitemia_nearest[i] <- df$Parasitemia[df$Date == time1]
  finch$IL_nearest[i] <- df$IL6_pg_ml[df$Date == time1]
  finch$Cum_Parasitemia_nearest[i] <- df$Cum_Parasitemia[df$Date == time1]
  finch$Survival[i] <- df$Survival[df$Date == time1]

  finch$Dif_Date[i] <- as.numeric(difftime(time1, finch$Date[i], units = "day"))
  
}



```

 

```{r}
df_final <- NULL

for(ring in unique(finch$Ring)) {
  
  df_predict <- finch %>% filter(Ring %in% ring)
  
  
  df_model <- finch_il %>% filter(Ring %in% ring) %>% filter(complete.cases(.))
  
  # Я пытался еще описывать данные полиномами, но по сути дела результаты не отличались.
  # degr = nrow(df_model) - 1
  # degr <- ifelse(degr > 5, 5, degr)
  # 
  # M_Parasitemia <- lm(Parasitemia ~ poly(DPI, degr), data = df_model)
  # df_predict$Parasitemia_M <- predict(M_Parasitemia, newdata = df_predict) 

  
  M_Parasitemia <- loess(Parasitemia ~ DPI, data = df_model, degree = 1)
  df_predict$Parasitemia_M <- predict(M_Parasitemia, newdata = df_predict) 

  M_Cum_Parasitemia <- loess(Cum_Parasitemia ~ DPI, data = df_model, degree = 1)
  df_predict$Cum_Parasitemia_M <- predict(M_Cum_Parasitemia, newdata = df_predict)

  M_IL <- loess(IL6_pg_ml ~ DPI, data = df_model, degree = 1)
  df_predict$IL_M <- predict(M_IL, newdata = df_predict)
  
  df_final <- rbind(df_final, df_predict)

}


df_final$Parasitemia_M[df_final$Parasitemia_M <0] <- 0
df_final$Cum_Parasitemia_M[df_final$Cum_Parasitemia_M <0] <- 0

df_final5 <- df_final

```

```{r}
finch_final <- merge(df_final5, rmr_paras_il)




```





```{r}

finch_final$Experiment <- factor(finch_final$Experiment)
finch_final$Ring <- factor(finch_final$Ring)
finch_final$Survival <- factor(finch_final$Survival)

finch_reduced <- finch_final %>% filter(Bad != 1) #удаляю из датасета строчки с неизмеренными значениями и строки с плохим измерением RMR

finch_reduced <- finch_reduced[complete.cases(finch_reduced), ]

```







## Динамика паразитемии

```{r}

finch_il_2 <- finch_il %>% filter(Experiment != "Control")

# M1_Par_gam <- gam(Parasitemia ~ s(DPI) + Experiment  + s(Ring, k = 53, bs = "re"),  method = "REML", data = finch_il_2, family = "nb")

M2_Par_gam <- gam(Parasitemia ~ s(DPI, by = Experiment) + Experiment  + s(Ring, k = 53, bs = "re"),  method = "REML", data = finch_il_2, family = "nb")


```



<!-- Проверка валидности -->
<!-- ```{r} -->
<!-- library(gratia) -->
<!-- appraise(M2_Par_gam) -->
<!-- ``` -->


```{r}
My_data <- expand.grid(Experiment = c("SGS1", "GRW2"), DPI = seq(0, max(finch_final$DPI)))

predicted <- predict(M2_Par_gam, newdata = My_data, exclude = "s(Ring)", newdata.guaranteed=TRUE, se.fit = TRUE, type = "response")

My_data$Fit <- predicted$fit
My_data$SE <- predicted$se.fit

Pl_parasetemia <-
ggplot(My_data, aes(x = DPI, y = Fit)) + 
  geom_line(aes(color = Experiment), size = 1) + 
  geom_ribbon(aes(ymin = Fit - 1.96*SE, ymax = Fit + 1.96*SE, group = Experiment), alpha = 0.2) +
  geom_point(data = finch_il_2, aes(y = Parasitemia, fill = Experiment), shape = 21, color = "black") +
  labs(y = "Parasitemia", color = "Lineage") + 
  scale_fill_manual(values = c("red", "blue")) +
  scale_color_manual(values = c("red", "blue")) +
  theme(legend.position = c(0.8, 0.9)) +
  guides(fill = "none")
  

```


```{r}
smooth_diff <- function(model, newdata, f1, f2, var, alpha = 0.05,
                        unconditional = FALSE) {
    xp <- predict(model, newdata = newdata, type = 'lpmatrix', exclude = "s(Ring)", newdata.guaranteed=TRUE)
    c1 <- grepl(f1, colnames(xp))
    c2 <- grepl(f2, colnames(xp))
    r1 <- newdata[[var]] == f1
    r2 <- newdata[[var]] == f2
    ## difference rows of xp for data from comparison
    X <- xp[r1, ] - xp[r2, ]
    ## zero out cols of X related to splines for other lochs
    X[, ! (c1 | c2)] <- 0
    ## zero out the parametric cols
    X[, !grepl('^s\\(', colnames(xp))] <- 0
    dif <- X %*% coef(model)
    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
    upr <- dif + (crit * se)
    lwr <- dif - (crit * se)
    data.frame(pair = paste(f1, f2, sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)
}


```




```{r}
comp1 <- smooth_diff(M2_Par_gam, My_data, 'SGS1', 'GRW2', 'Experiment', unconditional = TRUE)

comp <- cbind(DPI = seq(0, max(finch_final$DPI)),
              rbind(comp1))

signif_diff <- comp %>% filter((upper > 0 & lower > 0) | (upper < 0 & lower <0))



Pl_Parasitemia_difference <-
ggplot(comp, aes(x = DPI, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line() +
    facet_wrap(~ pair, ncol = 2)  +
    labs(x = "DPI", y = 'Smoothers difference') +  
    geom_hline(yintercept = 0, linetype = 2) + 
    geom_point(data = signif_diff, aes(x = DPI, y = 0), shape = 15, color = "black", fill = "black", size = 2)
  
```

```{r}
Pl_Fig1 <- plot_grid(Pl_parasetemia, Pl_Parasitemia_difference, labels = "AUTO")

Pl_Fig1
ggsave(filename = "Fig1.tiff", plot = Pl_Fig1, dpi = 500)
```

Figure +. Parasitemia at different days after inoculation (DPI) in two Plasmodium species. Panel A represents the observed number of merozoites (dots) and predicted GAM (lines, gray areas around the lines represent 95% CI). Panel B represents the difference between smoothers. Time periods when the difference between smoothers is significantly different from zero are marked by black rectangles.       






## Динамика RMR на фоне развития заболевания


```{r}

# M1_RMR_gam <- gam(log10(RMR) ~ s(DPI) + Experiment + log10(Mass)  + s(Ring, k = 53, bs = "re"),  method = "REML", data = finch_reduced)


M2_RMR_gam <- gam(log10(RMR) ~ s(DPI, by = Experiment) + Experiment + log10(Mass)  + s(Ring, k = 53, bs = "re"),  method = "REML", data = finch_reduced)



```






```{r}
My_data <- expand.grid(Experiment = levels(finch_final$Experiment), DPI = seq(0, max(finch_final$DPI)))
  # finch_final %>% select(Experiment, DPI) %>% unique(.)
My_data$Mass <- mean(finch_final$Mass)
  
predicted <- predict(M2_RMR_gam, newdata = My_data, exclude = "s(Ring)", newdata.guaranteed=TRUE, se.fit = TRUE)

My_data$Fit <- predicted$fit
My_data$SE <- predicted$se.fit

My_data2 <- expand.grid(Experiment = levels(finch_final$Experiment), DPI = seq(0, max(finch_final$DPI), 0.1))

My_data2$Fit [My_data2$Experiment != "Control"] <- predict(M2_Par_gam, newdata = My_data2[My_data2$Experiment != "Control", ], exclude = "s(Ring)", newdata.guaranteed=TRUE, type = "response") 

My_data2$Fit [My_data2$Experiment == "Control"] <- 0


Pl_RMR <-
ggplot(My_data, aes(x = DPI, y = 10^(Fit) )) + 
  geom_line(aes(color = Experiment), size = 1) + 
  geom_ribbon(aes(ymin = 10^(Fit - 1.96*SE), ymax = 10^(Fit + 1.96*SE)), alpha = 0.2) +
  geom_point(data = finch_final, aes(y = (RMR), fill = Experiment), shape = 21, color = "black") +
  labs(y = "RMR") +
  facet_wrap(~Experiment, ncol = 1) + 
  scale_color_manual(values = c("black", "red", "blue")) +
  scale_fill_manual(values = c("black", "red", "blue")) +
  guides(color = "none", fill = "none")



My_data_parasitemia <- expand.grid(RMR = seq(min(finch_final$RMR), max(finch_final$RMR), 1), DPI = seq(0,53,0.1), Experiment = c("SGS1", "GRW2"))

My_data_parasitemia2 <- expand.grid(RMR = seq(min(finch_final$RMR), max(finch_final$RMR), 1), DPI = seq(0,53,0.1), Experiment = c("Control"))

My_data_parasitemia2$Fit <- 0


My_data_parasitemia$Fit <- predict(M2_Par_gam, newdata = My_data_parasitemia, exclude = "s(Ring)", newdata.guaranteed=TRUE, type = "response" )
  
My_data_parasitemia <- rbind(My_data_parasitemia, My_data_parasitemia2)
  
  

Pl_RMR2 <-
ggplot(My_data, aes(x = DPI, y = 10^(Fit) )) + 
   geom_tile(data = My_data_parasitemia, aes(y = RMR, fill = Fit), alpha = 0.5) +
  geom_line(aes(color = Experiment), size = 1) + 
  geom_ribbon(aes(ymin = 10^(Fit - 1.96*SE), ymax = 10^(Fit + 1.96*SE)), alpha = 0.2) +
  geom_point(data = finch_final, aes(y = (RMR)), shape = 21, color = "black") +
  labs(y = "RMR") +
  facet_wrap(~Experiment, ncol = 1) + 
  scale_color_manual(values = c("black", "red", "blue")) +
  scale_fill_gradient(low = "white", high = "red") +
  guides(color = "none", fill = "none")


```



```{r}

comp1 <- smooth_diff(M2_RMR_gam, My_data, 'Control', 'SGS1', 'Experiment',unconditional = TRUE)

comp2 <- smooth_diff(M2_RMR_gam, My_data, 'Control', 'GRW2', 'Experiment',unconditional = TRUE)

comp3 <- smooth_diff(M2_RMR_gam, My_data, 'SGS1', 'GRW2', 'Experiment',unconditional = TRUE)

comp <- cbind(DPI = seq(0, max(finch_final$DPI)),
              rbind(comp1, comp2, comp3))

comp$pair <- factor(comp$pair, levels = c("Control-SGS1", "Control-GRW2", "SGS1-GRW2"))

signif_diff <- comp %>% filter((upper > 0 & lower > 0) | (upper < 0 & lower <0))


Pl_RMR_difference <-
ggplot(comp, aes(x = DPI, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line() +
    facet_wrap(~ pair, ncol = 1)  +
    labs(x = "DPI", y = 'Difference in RMR') +
    geom_hline(yintercept = 0, linetype = 2) + 
    geom_point(data = signif_diff, aes(x = DPI, y = 0), shape = 15, color = "black", fill = "black", size = 2)  

```

```{r}
Pl_Fig2_var1 <- plot_grid(Pl_RMR, Pl_RMR_difference, labels = "AUTO")

Pl_Fig2_var1

ggsave(filename = "Fig2_var1.tiff", plot = Pl_Fig2_var1, dpi = 500, width = 190, units = "mm")
```

Figure +. RMR at different days after inoculation (DPI) in different birds' groups.  Panel A represents the observed RMR level (dots) and GAM predictions (lines and 95% CI). Panel B represents the difference between smoothers. Time periods when the difference between smoothers is significantly different from zero are marked by black rectangles.       


### Второй вариант рисунка

```{r}
Pl_Fig2_var2 <- plot_grid(Pl_RMR2, Pl_RMR_difference, labels = "AUTO")

Pl_Fig2_var2

ggsave(filename = "Fig2_var2.tiff", plot = Pl_Fig2_var2, dpi = 500, width = 190, units = "mm")
```


Figure +. RMR at different days after inoculation (DPI) in different birds' groups.  Panel A represents the observed RMR level (dots) and GAM predictions (lines and 95% CI). Panel B represents the difference between smoothers. Time periods when the difference between smoothers is significantly different from zero are marked by black rectangles.       
The red shading reflects the level of parasitemia.


```{r}
cont_exp1 <- comp %>% filter(pair == "Control-SGS1") %>% filter(upper > 0 & lower > 0) %>% pull(DPI) %>% range()

cont_exp2 <- comp %>% filter(pair == "Control-GRW2") %>% filter(upper > 0 & lower > 0) %>% pull(DPI) %>% range()


exp1_exp2 <- comp %>% filter(pair == "SGS1-GRW2") %>% filter(upper < 0 & lower < 0) %>% pull(DPI) %>% range()
```


## Динамика интрелейкина

```{r}

# M1_IL_gam <- gam(log10(IL6_pg_ml) ~ s(DPI, k = 5) + Experiment  + s(Ring, k = 53, bs = "re"),  method = "REML", data = finch_il)

M2_IL_gam <- gam(log10(IL6_pg_ml) ~ s(DPI, by = Experiment, k = 5) + Experiment  + s(Ring, k = 53, bs = "re"),  method = "REML", data = finch_il)

```

```{r}
My_data <- expand.grid(Experiment = levels(finch_final$Experiment), DPI = seq(0, max(finch_final$DPI)))

predicted <- predict(M2_IL_gam, newdata = My_data, exclude = "s(Ring)", newdata.guaranteed=TRUE, se.fit = TRUE)

My_data$Fit <- predicted$fit
My_data$SE <- predicted$se.fit


Pl_IL <- 
ggplot(My_data, aes(x = DPI, y = 10^Fit)) + 
  geom_line(aes(color = Experiment), size = 1) + 
  geom_ribbon(aes(ymin = 10^(Fit - 1.96*SE), ymax = 10^(Fit + 1.96*SE)), alpha = 0.2) +
  geom_point(data = finch_il, aes(y = (IL6_pg_ml), fill = Experiment), shape = 21, color = "black") +
  labs(y = "IL-6 concentration") +
  facet_wrap(~Experiment, ncol = 1, scales = "free_y") +
  guides(color = "none") +
  scale_color_manual(values = c("black", "red", "blue")) + 
  scale_fill_manual(values = c("black", "red", "blue"))+
  guides(color = "none", fill = "none")
  


# My_data_parasitemia <- finch_final %>% 
#   group_by(Experiment) %>% 
#   do(expand.grid(IL = seq(min(.$IL), max(.$IL), 0.1), DPI = seq(0,53,0.1))) 
# 
# My_data_parasitemia2 <- My_data_parasitemia %>% filter(Experiment == "Control")
# My_data_parasitemia <- My_data_parasitemia %>% filter(Experiment != "Control")

My_data_parasitemia <- expand.grid(IL = seq(min(finch_final$IL), max(finch_final$IL), 0.1), DPI = seq(0,53,0.1), Experiment = c("SGS1", "GRW2"))

My_data_parasitemia2 <- expand.grid(IL = seq(min(finch_final$IL), max(finch_final$IL), 1), DPI = seq(0,53,0.1), Experiment = c("Control"))

My_data_parasitemia2$Fit <- 0


My_data_parasitemia$Fit <- predict(M2_Par_gam, newdata = My_data_parasitemia, exclude = "s(Ring)", newdata.guaranteed=TRUE, type = "response" )

My_data_parasitemia [My_data_parasitemia$Experiment == "GRW2" & My_data_parasitemia$IL > 30, ] <- NA  

My_data_parasitemia <- rbind(My_data_parasitemia, My_data_parasitemia2)

My_data_parasitemia <- My_data_parasitemia[complete.cases(My_data_parasitemia), ]
  


Pl_IL2 <- 
ggplot(My_data, aes(x = DPI, y = 10^Fit)) + 
 geom_tile(data = My_data_parasitemia, aes(y = IL, fill = Fit), alpha = 0.5) +
  geom_line(aes(color = Experiment), size = 1) + 
  geom_ribbon(aes(ymin = 10^(Fit - 1.96*SE), ymax = 10^(Fit + 1.96*SE)), alpha = 0.2) +
  geom_point(data = finch_il, aes(y = (IL6_pg_ml)), shape = 21, color = "black") +
  labs(y = "IL-6 concentration") +
  facet_wrap(~Experiment, ncol = 1, scales = "free_y") +
  guides(color = "none") +
  scale_color_manual(values = c("black", "red", "blue")) + 
  scale_fill_gradient(low = "white", high = "red")+
  guides(color = "none", fill = "none")


```


```{r}
comp1 <- smooth_diff(M2_IL_gam, My_data, 'Control', 'SGS1', 'Experiment', unconditional = TRUE)

comp2 <- smooth_diff(M2_IL_gam, My_data, 'Control', 'GRW2', 'Experiment', unconditional = TRUE)

comp3 <- smooth_diff(M2_IL_gam, My_data, 'SGS1', 'GRW2', 'Experiment', unconditional = TRUE)

comp <- cbind(DPI = seq(0, max(finch_final$DPI)),
              rbind(comp1, comp2, comp3))

comp$pair <- factor(comp$pair, levels = c("Control-SGS1", "Control-GRW2", "SGS1-GRW2"))

signif_diff <- comp %>% filter((upper > 0 & lower > 0) | (upper < 0 & lower <0))

Pl_IL_difference <-
ggplot(comp, aes(x = DPI, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line() +
    facet_wrap(~ pair, ncol = 1)  +
    labs(x = "DPI", y = 'Difference in IL-6 concentration') +
  geom_hline(yintercept = 0) +
      geom_hline(yintercept = 0, linetype = 2) + 
    geom_point(data = signif_diff, aes(x = DPI, y = 0), shape = 15, color = "black", fill = "black", size = 2)  

  
```


```{r}
Pl_Fig3_var1 <- plot_grid(Pl_IL, Pl_IL_difference, labels = "AUTO")

Pl_Fig3_var1

ggsave(filename = "Fig3_var1.tiff", plot = Pl_Fig3_var1, dpi = 500, width = 190, units = "mm")
```

Figure +. IL-6 at different days after inoculation (DPI) in different birds’ groups. Panel A represents the observed IL-6 concentration (dots) and GAM predictions (lines and 95% CI). Panel B represents the difference between smoothers. Time periods when the difference between smoothers is significantly different from zero are marked by black rectangles.




### Второй вариант рисунка


```{r}
Pl_Fig3_var2 <- plot_grid(Pl_IL2, Pl_IL_difference, labels = "AUTO")

Pl_Fig3_var2

ggsave(filename = "Fig3_var2.tiff", plot = Pl_Fig3_var2, dpi = 500, width = 190, units = "mm")

```


Figure +. IL-6 at different days after inoculation (DPI) in different birds’ groups. Panel A represents the observed IL-6 concentration (dots) and GAM predictions (lines and 95% CI). Panel B represents the difference between smoothers. Time periods when the difference between smoothers is significantly different from zero are marked by black rectangles. 
The red shading reflects the level of parasitemia.


