---
title: "Predators regulate the taxonomic structure of mixed *Mytilus edulis* and *M.trossulus* settlements in the shallow waters of the White Sea"
author: "Vadim Khaitov, Anna Makarycheva, Renata Nematova, Alena Evdokimova"
date: ""
output:
  word_document:
    reference_docx: article_template.docx
    fig_width: 9
    fig_height: 9

---

```{r setup, include=FALSE, echo=FALSE}
library(knitr)

opts_chunk$set(echo = FALSE, cache = FALSE, fig.align ="center", warning = FALSE, message = FALSE)

library(ggplot2)
library(ggrepel)
library(cowplot)
library(readxl)

library(dplyr)

library(ggmap)
library(mapproj)
library(maps)
library(rgeos) #этот пакет содержит какую-то хрень, которая позволяет обойти проблему пр чтении фалов средсвами maptools. 
#Att! Этот пакет должен быть загружен до maptools

library(mapdata)
library(maptools) # Rgshhs
library(PBSmapping)
library(gridExtra)
library(grid)


library(lme4)
library(mgcv)
library(broom.mixed)
library(broom)


```



Abstract. 

++++++


## Introduction

Mussels of several cryptic species, known as “*Mytilus edulis*” complex (++, Riginos and Cunningham, 2005), are widely distributed around North hemisphere. Species of this complex frequently coexist sympatricaly as do, for instance, *M. edulis* (Me) and *M. trossulus* (Mt) along both coasts of the North Atlantic (Riginos and Cunningham, 2005; Väinölä and Strelkov, 2011 and references therein). The proportion of both species in mixed populations vary in broad limits (Katolikova et al. +++) but factors regulating species composition in locality is poorly understood. 



<!-- ```{r ft.split=FALSE} -->
<!-- library(flextable) -->
<!-- data<-regulartable(head(iris, 10)) -->
<!-- data -->
<!-- ``` -->



```{r}
library(flextable)
# if(require("equatags") && mathjax_available()){

eqs <- c(
  "P_{Mt}",
  "a \\ne 0",
  "x = {-b \\pm \\sqrt{b^2-4ac} \\over 2a}")
df <- data.frame(formula = eqs)
# df


ft <- flextable(df)
ft <- compose(
  x = ft, j = "formula",
  value = as_paragraph(as_equation(formula, width = 2, height = .5)))
ft <- align(ft, align = "center", part = "all")
ft <- width(ft, width = 2)
ft
# }
```


